Practical Considerations for Lomb-Scargle Periodic Analysis

### 2016-12-07 paper.tex
Add some figures & rearrange the outline

The Lomb-Scargle periodogram \citep{Lomb76, Scargle82}
is a well-known algorithm for detecting periodicity
in unevenly-sampled time-series, particularly within the astronomy community.
For example, consider the data showing in \Fig{LINEAR_data}.
This is an irregularly-sampled timeseries showing a single object from the
LINEAR survey, with magnitude measured 280 times over the course of five and
a half years.

It is clear that this object varies with an amplitude of around 0.6 magnitudes.
Computing the Lomb-Scargle periodogram for the data gives us a measure of the
power as a function of period of oscillation (\fig{LINEAR_power}, left)\footnote{
This periodogram was computed using tools in the AstroPy project \citep{Astropy2013};
code is available at \url{http://github.com/jakevdp/PracticalLombScargle/}
}, from
which we can determine the period of oscillation of approximately 2.58 hours.
The right panel of \fig{LINEAR_power} shows a folded visualization of
the same data as \fig{LINEAR_data} -- i.e.{} plotted as a function of phase
rather than time.

Often this is exactly how the Lomb-Scargle periodogram is presented: as a clean,
well-defined procedure to detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be considered
when applying a Lomb-Scargle analysis to real-world datasets, and I have found
that these are rarely presented together at an introductory level.
This paper seeks to fill that gap, and provide a practical, semi-technical
guide to the effective use of the Lomb-Scargle method of periodic analysis.

\section{Background: Continuous Fourier Transform}

To understand what the Lomb-Scargle periodogram is doing, we will first step
back and review the subject of Fourier analysis of continuous signals.
If we have a continuously-defined signal $g(t)$, we can compute its Fourier
transform $\hat{g}(f)$ using the following integral:
\begin{equation}
    \hat{g}(f) \equiv \int_{-\infty}^\infty g(t) e^{-2\pi i f t} dt
    \eqlabel{FT-def}
\end{equation}
The inverse relationship is given by:
\begin{equation}
    g(t) \equiv \int_{-\infty}^\infty \hat{g}(f) e^{+2\pi i f t} df
    \eqlabel{IFT-def}
\end{equation}
This can be straightforwardly confirmed using the linearity of integration and
the integral definition of the Dirac delta function.
For convenience below, we will define the Fourier transform operator $\mathcal{F}$,
such that
\begin{eqnarray}
    \mathcal{F}\{g\} &=& \hat{g} \\
    \mathcal{F}^{-1}\{\hat{g}\} &=& g
\end{eqnarray}

@@

\begin{description}
\item[The Fourier Transform of a Sine wave is a delta function at its frequency.]
    That is, if $g(t) = \sin(2\pi f_0 t)$, then
    \begin{equation}
        \hat{g}(f) \propto \delta(f - f_0) - \delta(f + f_0)
    \end{equation}
\item[The Fourier Transform is a Linear Operation.]
    This means that for functions $a$ and $b$ and scalar $A$,
    \begin{eqnarray}
        \mathcal{F}(a + b) &=& \mathcal{F}(a) + \mathcal{F}(b) \\
        \mathcal{F}(A \cdot a) &=& A\cdot\mathcal{F}(a).
    \end{eqnarray}
    These properties follow straightforwardly from the linearity of integration.
\item[The Fourier Transform of a time-shifted function is unchanged up to a phase.]
    That is, if $g_\tau(t) = g(t - \tau)$, then
    \begin{equation}
        \hat{g}_\tau(f) = e^{-2\pi i f \tau}\hat{g}(f)
    \end{equation}
    This can be confirmed via a change of variables in \eq{FT-def}.
\end{description}

Taken together, what this tells us is that if we define the spectral density
\begin{equation}
    \mathcal{P}(g) \equiv \left|\mathcal{F}(g)\right|^2
\end{equation}
we have a quantity that is {\it insensitive to time-shifts} and consists of a
{\it delta function for each sinusoidal frequency within the data}. In other
words, the spectral density {\it P} measures ...

Because of the first two properties, we see that the Fourier transform can essentially
extract sinusoidal components within a signal:

\begin{description}

\item[Convolution Theorem.]
    The convolution theorem states that the
    Fourier transform of a convolution is a product of the Fourier transforms:
    \begin{equation}
      \mathcal{F}\{g \ast h\} = \mathcal{F}\{g\}\mathcal{F}\{h\}
    \end{equation}
    where the convlution operator $\ast$ is defined by
    \begin{equation}
      \{g \ast h\}(t) \equiv \int_{-\infty}^\infty g(t - \tau) h(\tau) d\tau.
    \end{equation}
    This can be confirmed via the linearity of integration and the integral
    definition of the Dirac delta function.

\item[The Fourier Transform of a Gaussian is a Gaussian.]
    The Gaussian function is defined as
    \begin{equation}
    G(t;\sigma) \propto \exp\left[\frac{-t^2}{2\sigma^2}\right].
    \end{equation}
    The Fourier transform of a Gaussian is another Gaussian:
    \begin{equation}
    \mathcal{F}\{G(t;\sigma)\} \propto G(f; \sigma_f)
    \end{equation}
    where $\sigma_f = (2\pi\sigma)^{-1}$. This can be shown by completing the
    square within the integrand.
\end{description}

\subsubsection{The Convolution Theorem}

The Fourier transform is central to an identity known as {\it the convolution theorem}.
The convolution of two functions $g(t)$ and $h(t)$ is itself a function of $t$, denoted $\{g \ast h\}(t)$, and is defined as

The convolution theorem states that the Fourier transform of a convolution is equal to the product of Fourier transforms; that is,

This property can also be shown (relatively) straightforwardly from the convolution and Fourier transform definitions.

\subsubsection{$\mathcal{F}$ of a Gaussian is a Gaussian}
One of the special functions under a Fourier transform is the Gaussian distribution:

\begin{equation}
    g(t; \mu, \sigma) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left[\frac{- (x - \mu)^2}{2\sigma^2}\right]
\end{equation}, which is proportional to

\subsubsection{$\mathcal{F}$ of a Periodic Function is a Delta Function}

A particularly special function in the complex exponential, $e^{2\pi i\nu_0 t} \equiv \cos(2\pi \nu_0 t) + i\sin(2\pi\nu_0 t)$, which has the Fourier transform $\mathcal{F}\{e^{2\pi i\nu_0 t}\} = \delta(\nu - \nu_0)$, which follows directly from the definition of the Dirac delta function:

\begin{equation}
  \delta(\nu) \equiv \int_{-\infty}^\infty e^{-2\pi i\nu t}dt
\end{equation}

Similarly, if we have a signal composed of a sum of complex exponentials, the linearity of the Fourier transform assures that the transform is simply a sum of delta functions:
\begin{equation}
  \mathcal{F}\left\{ \sum_{n=1}^N e^{-2\pi i \nu_n t} \right\} = \sum_{n=1}^N \delta(\nu - \nu_n)
\end{equation}
This result shows the sense in which a Fourier transform is an ideal measure of periodicity: the Fourier transform pulls-out and identifies the individual periodic components of any input.

@@

\begin{itemize}
\item convolution function and windowing
\end{itemize}

@@

\begin{itemize}
\item Talk about peak width and False Alarm Probability; recommend bootstrap (show example of when analytic FAP fails?)
\end{itemize}

@@

\begin{itemize}
\item Talk about Press \& Rybicki method; show some benchmarks.
\end{itemize}

>>>>

The Lomb-Scargle periodogram \citep{Lomb76, Scargle82}
is a well-known algorithm for detecting periodicity
in unevenly-sampled time-series, particularly within the astronomy community.
For example, consider the data shown in \fig{LINEAR_data}.
This is an irregularly-sampled timeseries showing a single object from the
LINEAR survey, with magnitude measured 280 times over the course of five and
a half years.

By eye, it is clear that the brightness of the object varies in time with a range spanning approximately 0.8 magnitudes, but what is not immediately clear is that this variation is periodic in time.
The Lomb-Scargle periodogram is a method that allows efficient computation of a Fourier-like power spectrum from such unevenly-sampled data, resulting in an intuitive means of determining the period of oscillation.

The left panel of \fig{LINEAR_power} shows the Lomb-Scargle periodogram computed for the data in \fig{LINEAR_data}
Computing the Lomb-Scargle periodogram for the data gives us a measure of the
power as a function of period of oscillation (\fig{LINEAR_power}, left), from
which we can determine the period of oscillation of approximately 2.58 hours.
The right panel of \fig{LINEAR_power} shows a folded visualization of
the same data as \fig{LINEAR_data} -- i.e.{} plotted as a function of phase
rather than time.\footnote{
    The periodogram in \fig{LINEAR_power}was computed using
    {\tt astropy.stats.LombScargle}
    from the AstroPy project \citep{Astropy2013};
    Python code to reproduce this figure is available at
    \url{http://github.com/jakevdp/PracticalLombScargle/}
}

Often this is exactly how the Lomb-Scargle periodogram is presented: as a clean, well-defined procedure to detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be considered when applying a Lomb-Scargle analysis to real-world datasets, and I have found that these are rarely presented together at an introductory level.
This paper seeks to fill that gap, and provide a practical, semi-technical guide to the effective use of the Lomb-Scargle method of periodic analysis.
This paper does not seek a complete or rigorous treatment of the mathematics involved, but rather seeks to develop the intuition of {\it how} to think about the issues involved.

\section{Background: the Continuous Fourier Transform}

In order to understand how we should interpret the Lomb-Scargle periodogram, we will first briefly step back and review the subject of Fourier analysis of continuous signals.
Consider a continuously-defined signal $g(t)$.
Its Fourier transform is given by the following integral:
\begin{equation}
    \hat{g}(f) \equiv \int_{-\infty}^\infty g(t) e^{-2\pi i f t} dt
    \eqlabel{FT-def}
\end{equation}
The inverse relationship is given by:
\begin{equation}
    g(t) \equiv \int_{-\infty}^\infty \hat{g}(f) e^{+2\pi i f t} df
    \eqlabel{IFT-def}
\end{equation}
For convenience we will also define the Fourier transform operator
$\mathcal{F}$, such that
\begin{eqnarray}
    \mathcal{F}\{g\} &=& \hat{g} \\
    \mathcal{F}^{-1}\{\hat{g}\} &=& g
\end{eqnarray}

The continuous Fourier transform has a number of useful properties that we will make use of in our discussion.

\begin{description}
   \item[The Fourier transform is a linear operation.]
     That is, for any constant $A$ and any functions $f$ and $g$, we can write
     \begin{eqnarray}
       \mathcal{F}\{f(t) + g(t)\} &=& \mathcal{F}\{f(t)\} + \mathcal{F}\{g(t)\}\nonumber\\
       \mathcal{F}\{A f(t)\} &=& A\mathcal{F}\{f(t)\},
     \end{eqnarray}
     which follows from the linearity of the Fourier integral.

   \item[The Fourier transform of sinusoid with frequency $f_0$ is a sum of delta functions at $\pm f_0$.]
     Given the integral definition of the Dirac delta function, it is straightforward to show that
     \begin{equation}
       \mathcal{F}\{e^{2\pi f_0 t}\} = \delta(f - f_0).
       \eqlabel{delta_FT}
     \end{equation}
     Euler's formula shows how such a complex exponential
     can be expressed in terms of sines and cosines:
     \begin{equation}
       e^{2\pi f_0 t} = \cos(2\pi f_0 t) + i\sin(2\pi f_0 t).
       \eqlabel{Euler_formula}
     \end{equation}
     Using the linearity of the Fourier transform, we can take these identites
     and derive the following:
     \begin{eqnarray}
       \mathcal{F}\{\cos(2\pi f_0 t)\} &=& \frac{1}{2}\left[\delta(f - f_0) + \delta(f + f_0)\right]\\
       \mathcal{F}\{\sin(2\pi f_0 t)\} &=& \frac{1}{2i}\left[\delta(f - f_0) - \delta(f + f_0)\right].
     \end{eqnarray}
     In other words, a sinusoidal signal with frequency $f_0$ has a Fourier transform consisting of a weighted sum of delta functions at $\pm f_0$.

   \item[A time-shift imparts a phase in the Fourier transform.]
     That is for any function $g(t)$,
     \begin{equation}
       \mathcal{F}\{g(t - t_0)\} = e^{-2\pi i ft_0}\mathcal{F}\{g(t)\}
     \end{equation}
     The key observation here is that the time-shift does not change the
     amplitude of the result, but only the phase.
\end{description}

There are a couple useful consequences of these properties.

...

\subsection{Some Useful Fourier Pairs}

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{fig03_Fourier_pairs}
\caption{Visualization of important Fourier pairs.}
\end{figure}

Dirac comb
\begin{equation}
    C(t, \Delta t) = \sum_{k=-\infty}^\infty \delta(t - k\Delta t)
\end{equation}
Fourier transform is
\begin{equation}
    \mathcal{F}\{C(t, \Delta t)\} = \frac{1}{\Delta t} C(f, 1/\Delta t)
\end{equation}

Tophat and Sinc
\begin{equation}
  tophat(t; T) \equiv \left\{
  \begin{array}{ll}
    1 / T & {\rm if}\ |t| < T / 2 \\
    0     & {\rm otherwise}
  \end{array}
  \right.
\end{equation}
Fourier transform is
\begin{equation}
  sinc(t) \equiv \frac{\sin(\pi t)}{\pi t}
\end{equation}

Gaussian

\begin{itemize}
  \item Chart of Fourier pairs
\end{itemize}

\subsection{Windowing and Convolution}

\begin{itemize}
  \item mention convolution theorem; show how it leads to windowing
  \item figure: show how windowing works
\end{itemize}

\section{From Continuous to Discrete Signals}

\begin{itemize}
  \item discrete signal as a Dirac comb window
  \item derive the Nyquist limit from the Dirac comb FT
  \item derive the Schuster periodogram
  \item mention the FFT
\end{itemize}

\section{Non-uniform sampling}

\begin{itemize}
  \item nonuniform signal as sum-of-deltas window function
  \item Nyquist limit does not exist!
  \item 
\end{itemize}

\section{The Lomb-Scargle Periodogram}

\begin{itemize}
  \item Show relation to Schuster periodogram
  \item Show relation to least squares fitting of sinusoid
  \item Key: LS measures {\it windowed} power spectrum
  \item Show effects of truely random sampling; of Kepler-like sampling; of ground-based nightly sampling
\end{itemize}

\section{Using Lomb-Scargle Effectively}

\begin{itemize}
  \item Frequency grid
  \item Peak width
\end{itemize}

\subsection{Uncertainties in Periods}

\begin{itemize}
  \item Peak Width
  \item False Alarm Probability and why it fails
  \item Baluev method and why it's not enough
  \item Boostrap

\end{itemize}

\begin{itemize}
\item Press \& Rybicki method; show some benchmarks.
\item NFFT method
\end{itemize}

\begin{itemize}
\item Bayesian view doesn't make sense
\item Multiterm makes the true period fit better, but also bumps the background noise.
\end{itemize}

\newpage

\section{Draft stuff below...}

\begin{itemize}
    \item transform and inverse
    \item properties (delta function $\to$ sinusoid, time shift $\to$ phase, convolution theorem)
    \item motivating the power spectrum
    \item some important Fourier pairs (sinusoid$\to$delta, dirac comb$\to$dirac comb, Gaussian$\to$gaussian, top-hat$\to$sinc)
\end{itemize}

@@

\begin{description}
\item[The Fourier Transform is a Linear Operation.]
This means that for functions $a$ and $b$ and scalar $A$,
\begin{eqnarray}
    \mathcal{F}(a + b) &=& \mathcal{F}(a) + \mathcal{F}(b) \\
    \mathcal{F}(A \cdot a) &=& A\cdot\mathcal{F}(a).
\end{eqnarray}
These properties follow straightforwardly from the linearity of integration.
\item[The Fourier Transform of a Sinusoid with frequency $f_0$ is a pair of delta
      functions at $\pm f_0$.]
    That is, if $g(t) = \sin(2\pi f_0 t)$, then
    \begin{equation}
        \hat{g}(f) = \frac{i}{2} \bigg(\delta(f - f_0) - \delta(f + f_0) \bigg)
    \end{equation}
    This can be shown straightforwardly using Euler's formula and the integral
    definition of the Dirac delta function.
\item[The Fourier Transform of a time-shifted function is unchanged up to a phase.]
    That is, if $g_\tau(t) = g(t - \tau)$, then
    \begin{equation}
        \hat{g}_\tau(f) = \phi(\tau)\hat{g}(f)
    \end{equation}
    where $\phi$ is a phase that depends on the time-shift $\tau$.
    This can be confirmed via a change of variables in \eq{FT-def}.
\end{description}
These properties are central to the well-known ability of the Fourier transform
to measure frequency information from a continuous signal: a signal
with a periodic component at frequency $f_0$ will have a Fourier transform
reflecting that frequency. Moreover, we can remove the effect of the phase by
taking the squared magnitude of the Fourier transform; this is the power
spectral density:
\begin{equation}
    \mathcal{P}(g) \equiv \left|\mathcal{F}(g)\right|^2
\end{equation}
Because this operation removes the phase, we now have a quantity that is
{\it insensitive to time-shifts} and consists of a
{\it delta function for each sinusoidal frequency within the data}.
If we want to detect periodic components in a continuous signal,
this will be a very useful quantity.

\subsection{Other Identities}

\begin{description}
\item[Convolution Theorem.]
    The convolution theorem states that the
    Fourier transform of a convolution is a product of the Fourier transforms:
    \begin{equation}
      \mathcal{F}\{g \ast h\} = \mathcal{F}\{g\}\mathcal{F}\{h\}
    \end{equation}
    where the convlution operator $\ast$ is defined by
    \begin{equation}
      \{g \ast h\}(t) \equiv \int_{-\infty}^\infty g(t - \tau) h(\tau) d\tau.
    \end{equation}
    This can be confirmed via the linearity of integration and the integral
    definition of the Dirac delta function.
\item[The Fourier Transform of a Gaussian is a Gaussian.]
    The Gaussian function is defined as
    \begin{equation}
    G(t;\sigma) \propto \exp\left[\frac{-t^2}{2\sigma^2}\right].
    \end{equation}
    The Fourier transform of a Gaussian is another Gaussian:
    \begin{equation}
    \mathcal{F}\{G(t;\sigma)\} \propto G(f; \sigma_f)
    \end{equation}
    where $\sigma_f = (2\pi\sigma)^{-1}$. This can be shown by completing the
    square within the integrand.
\end{description}

@@

\begin{itemize}
\item convolution theorem and windowing
\item infinite, regularly-sampled observations
\item finite, regularly-sampled observations
\item finite, irregularly-sampled observations
\end{itemize}

@@

\begin{itemize}
\item Peak width
\item False Alarm Probability and why it fails
\item Baluev method and why it's not enough
\item Boostrap
\end{itemize}

@@

\begin{itemize}
\item Talk about Press \& Rybicki method; show some benchmarks.
\item NFFT method
\end{itemize}
